---
title: "nf-core/metatdenovo - kimchi dataset analysis"
author: "danilo.dileo@lnu.se, emelie.nilsson@lnu.se, ..., daniel.lundin@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
format:
  html:
    code-fold: true
    toc: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
bibliography:
  - bibliography.bib
  - grateful-refs.bib
---

```{r setup}
#| label: setup
#| echo: false
#| cache: false

knitr::opts_chunk$set(echo = TRUE, fig.path='figures/', cache = TRUE, fig.width = 10)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries}
#| label: libraries
#| message: false
#| cache: false
#| include: false

library(readr)
library(forcats)
library(tibble)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(lubridate)
library(purrr)
library(stringr)
library(ggplot2)
library(kfigr)
library(knitr)
library(DT)
library(grateful)
library(patchwork)
```

```{r constants}
#| label: constants
TAXON_RANKS_SORTED = c('root', 'domain', 'phylum', 'class', 'order', 'family', 'genus', 'species')

k.rank.taxa <- c("Leuconostoc_carnosum", "Leuconostoc_gelidum",
                       "Leuconostoc_inhae", "Weissella_koreensis",
                       "Latilactobacillus_sakei", "Leuconostoc_mesenteroides")
k.rank.p1 <- c("Leuconostoc_carnosum", "Leuconostoc_gelidum",
                       "Leuconostoc_inhae", "Weissella_koreensis",
                       "Latilactobacillus_sakei", "Leuconostoc_mesenteroides", "Others")
k.color.taxon <- c(
  "#999999",   # grey
  "#E69F00",   # orange
  "#56B4E9",   # sky blue
  "#009E73",   # teal green
  "#F0E442",   # soft yellow
  "#0072B2",   # dark blue
  "#D55E00",   # reddish brown
  "#CC79A7",   # purple-pink
  "#332288",   # navy
  "#88CCEE",   # light blue
  "#44AA99"    # green-cyan
)

k.rank.order <- c("Bacteria", "Gammaproteobacteria",
                       "Lachnospirales", "Plants",
                       "Oscillospirales", "Lactobacillales", 'others', 'UBA9160', "Viruses", "Xanthomonadales")
k.rank.sample = c("J7", "J13", "J18", "J25", "J29")
k.rank.genes = c("scrK", "uxuB", "actk", "xpkA", "gntZ",
                "G6p1dh", "lctd", "pyk", "eno", "pgm",
                "pgk", "gap", "tpiS","fba", "pfkA", "pgi","glK")
```

```{r read-execution-traces}
#| label: read-data
#| cache-lazy: false
#| warning: false

extraces <- Sys.glob('nextflow/kimchi/*/pipeline_info/execution_trace*') %>%
  read_tsv(show_col_types = FALSE, id = 'fname') %>%
  mutate(
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*'),
    p       = str_remove(name, '.*:'),
    process = str_remove(p, ' .*'),
    sample  = str_replace(p, '.*\\((.*)\\)', '\\1'),
    durationc  = str_remove(duration, ' *\\d+ms') %>%
      ifelse(str_detect(., '\\dh'), ., str_c('0h', ., sep = ' ')) %>%
      ifelse(str_detect(., '\\dm'), ., str_c('0m', ., sep = ' ')) %>%
      ifelse(str_detect(., '\\ds'), ., str_c(., '0s', sep = ' ')) %>%
      hms(),
    realtimec  = str_remove(realtime, ' *\\d+ms') %>%
      ifelse(str_detect(., '\\dh'), ., str_c('0h', ., sep = ' ')) %>%
      ifelse(str_detect(., '\\dm'), ., str_c('0m', ., sep = ' ')) %>%
      ifelse(str_detect(., '\\ds'), ., str_c(., '0s', sep = ' ')) %>%
      hms(),
    num_cpus   = ( str_remove(`%cpu`, '%') %>%
      as.double()/100 ) %>%
      ceiling(),
    peak_rssc = case_when(
      str_detect(peak_rss, 'GB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^3,
      str_detect(peak_rss, 'MB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^2,
      str_detect(peak_rss, 'KB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^1,
      str_detect(peak_rss, 'B')  ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^0,
      TRUE                       ~ 0
    ),
    peak_vmemc = case_when(
      str_detect(peak_vmem, 'GB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^3,
      str_detect(peak_vmem, 'MB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^2,
      str_detect(peak_vmem, 'KB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^1,
      str_detect(peak_vmem, 'B')  ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^0,
      TRUE                       ~ 0
    ),
    rcharc = case_when(
      str_detect(rchar, 'GB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^3,
      str_detect(rchar, 'MB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^2,
      str_detect(rchar, 'KB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^1,
      str_detect(rchar, 'B')  ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^0,
      TRUE                       ~ 0
    ),
    wcharc = case_when(
      str_detect(wchar, 'GB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^3,
      str_detect(wchar, 'MB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^2,
      str_detect(wchar, 'KB') ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^1,
      str_detect(wchar, 'B')  ~ str_remove(peak_rss, ' .*') %>% as.numeric() * 1024^0,
      TRUE                       ~ 0
    )
  ) %>%
  select(-p, -fname) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  relocate(project, run, assembly, assembler, orf_caller, bbnorm, minlen, process, sample)
```

```{r read-overall-stats}
#| label: read-overall-stats
#| warning: false
#| cache-lazy: false

k.overall_stats <- Sys.glob('nextflow/kimchi/*/summary_tables/*overall_stats.tsv.gz') %>%
  read_tsv(id = 'fname', show_col_types = FALSE) %>%
  mutate(
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*')
  ) %>%
  select(-fname) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  relocate(project, run, assembly, assembler, orf_caller, bbnorm, minlen) %>%
  mutate( sample = str_replace(sample, "SRR443366", "J7"),
          sample = str_replace(sample, "SRR443367", "J13"),
          sample = str_replace(sample, "SRR443368", "J18"),
          sample = str_replace(sample, "SRR443369", "J25"),
          sample = str_replace(sample, "SRR443370", "J29"))

# pubmed search for intro plot
pm <-  read_delim('tables/pubmed_metat.csv', delim = ';', show_col_types = FALSE)
```

```{r read-transrate}
#| label: read-transrate
#| warning: false
#| cache-lazy: false

transrates <- Sys.glob('nextflow/kimchi/*/transrate/*.csv') %>%
  read_csv(id = 'fname', show_col_types = FALSE) %>%
  mutate(
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*')
  ) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  select(-fname, -run, -orf_caller) %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  relocate(project, assembly, assembler, bbnorm, minlen) %>%
  distinct()
```

```{r read-counts}
#| label: read-counts
#| cache-lazy: false
#| warning: false

counts <- Sys.glob('nextflow/kimchi/*/summary_tables/*.counts.tsv.gz') %>%
  read_tsv(id = 'fname', show_col_types = FALSE) %>%
  mutate(
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*')
  ) %>%
  select(-fname) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  relocate(project, run, assembly, assembler, orf_caller, bbnorm, minlen, sample)
```

```{r read-eukulele}
#| message: false
#| include: false
#| label: read-eukulele
#| cache-lazy: false
#| warning: false
# it looks like Eukulele failed to annotate the genes. could it be an error of the new version?
eukulele <- Sys.glob('nextflow/kimchi/*/summary_tables/*_taxonomy.tsv.gz') %>%
  read_tsv(id = 'fname', show_col_types = FALSE) %>%
  mutate(
    db      = str_replace(fname, 'nextflow\\/.*\\.(.*)_taxonomy.tsv.gz', '\\1'),
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*')
  ) %>%
  select(-fname) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  relocate(project, run, assembly, assembler, orf_caller, bbnorm, minlen, db) %>%
  mutate(
    taxon_rank = case_when(
      str_detect(domain,  'uncl\\.')  ~ 'root',
      str_detect(phylum,  'uncl\\.')  ~ 'domain',
      str_detect(class,   'uncl\\.')  ~ 'phylum',
      str_detect(order,   'uncl\\.')  ~ 'class',
      str_detect(family,  'uncl\\.')  ~ 'order',
      str_detect(genus,   'uncl\\.')  ~ 'family',
      str_detect(species, 'uncl\\.')  ~ 'genus',
      TRUE                            ~ 'species'
    ) %>% factor(levels = TAXON_RANKS_SORTED, ordered = TRUE)
  )

k.taxonomy.euk <- read.table("nextflow/kimchi/megahit.prokka.without_bbnorm/summary_tables/kimchi-phylodb.out", sep = "\t", header = TRUE, fill = TRUE) %>%
  select(-1) %>%
  rename(orf = transcript_name) %>%
  group_by(orf) %>%
  filter(max_pid == max(max_pid)) %>%
  ungroup() %>%
  separate(
    full_classification,
    c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";"
  ) %>%
  mutate(
    domain  = ifelse(is.na(domain)  | domain  == '', 'Uncl.', domain),
    phylum   = ifelse(is.na(phylum)   | phylum   == '', sprintf("%s uncl.", domain), phylum),
    class   = ifelse(is.na(class)   | class   == '', sprintf("%s uncl.", phylum), class),
    order   = ifelse(is.na(order)   | order   == '', sprintf("%s uncl.", class),  order),
    family  = ifelse(is.na(family)  | family  == '', sprintf("%s uncl.", order),  family),
    genus   = ifelse(is.na(genus)   | genus   == '', sprintf("%s uncl.", family), genus),
    species = ifelse(is.na(species) | species == '', sprintf("%s uncl.", genus),  species)
  ) 
```

```{r read-eggnogs}
#| label: read-eggnogs
#| cache-lazy: false

k.eggnogs <- Sys.glob('nextflow/kimchi/*/summary_tables/*.emapper.tsv.gz') %>%
  read_tsv(id = 'fname', show_col_types = FALSE) %>%
  mutate(
    project = str_replace(fname, 'nextflow\\/([^/]+)\\/.*', '\\1'),
    run     = str_remove(fname, 'nextflow\\/[^/]+\\/') %>% str_remove('\\/.*')
  ) %>%
  select(-fname) %>%
  separate(run, c('assembler', 'orf_caller', 'bbnorm', 'minlen'), remove = FALSE, sep = '\\.', fill = 'right') %>%
  mutate(minlen = ifelse(is.na(minlen), 0, str_remove(minlen, 'minlen') %>% as.integer())) %>%
  mutate(assembly = sprintf("%s%s%s", assembler, ifelse(bbnorm == 'with_bbnorm', '*', ''), ifelse(minlen > 0, '†', ''))) %>%
  rename_with(str_to_lower) %>%
  relocate(project, run, assembly, assembler, orf_caller, bbnorm, minlen)
```

```{r read-prokkas}
#| label: read-prokkas
#| cache-lazy: false
k.prokkas <- read_tsv('nextflow/kimchi/megahit.prokka.without_bbnorm/summary_tables/megahit.prokka.prokka-annotations_kimchi.tsv.gz',
                      show_col_types = FALSE, guess_max = 1e6)

```

```{r tax-annots-og}
#| label: tax-annots-og
#| cache-lazy: false

original.tax <- tibble(
  sample = c(rep("Lca",7), rep("Lcc",7), rep("Hca",7), rep("Hcc",7)),
  family = c(rep(c("Other flavo.", "Flavobacteriaceae", "Other gammapr.", "Alteromonadaceae", "Other alphapr.", "SAR11 clade", "Rhodobacteraceae"),4)),
  #family = c(rep(c("Schleiferiaceae", "Flavobacteriaceae", "Other gammapr.", "Alteromonadaceae", "Thalassospiraceae", "Pelagibacteraceae", "Rhodobacteraceae"),4)),
  tpm = c(0.1, 0.5, 2.3, 2.4, 0.3, 0.4, 0.1,
          0.1, 0.5, 2.3, 2.3, 0.5, 0.2, 0.1,
          0.3, 0.5, 2.5, 0.7, 0.4, 0.1, 0.1,
          0.1, 0.7, 2.5, 1.7, 0.9, 0.2, 0.2)
  )
```

```{r func-annots-og}
#| label: func-annots-og
#| cache-lazy: false

original.fun <- tibble(
  sample = c(rep("Lca",13), rep("Lcc",13), rep("Hca",13), rep("Hcc",13)),
  SEED = c(rep(c("Protein metabolism", "Clustering-based subsystem", "RNA Metabolism", "Membrane Transport", "Carbohydrates", "Miscellaneous", "Amino Acids and Derivates", "Respiration", "Cofactors, vitamins, Pigments", "Iron acquisition and mtabolism", "Nucleosiles and Nucleotides", "Fatty Acids", "Photosynthesis"),4)),
  tpm = c(2.9, 1.8, 0.9, 1.1, 0.8, 0.88, 0.70, 0.88, 0.5, 0.5, 0.3, 0.15, 0.35,
          3.3, 2, 1, 1, 0.82, 0.89, 0.72, 0.72, 0.52, 0.48, 0.33, 0.17, 0.3,
          2.9, 2.4, 0.85, 0.85, 0.96, 0.78, 0.80, 0.70, 0.57, 0.45, 0.4, 0.25, 0.15,
          3.1, 2.2, 0.88, 0.88, 0.97, 0.88, 0.78, 0.77, 0.56, 0.46, 0.35, 0.3, 0.2))
```

# Introduction

Cite nf-core [@nf-core].


# Materials and Methods

## R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in Table @tbl-cite-packages.

```{r tbl-cite-packages}
#| label: tbl-cite-packages
#| cache: false
#| tbl-cap: Versions of R and packages used in this analysis.

cite_packages(output = "table", pkgs = "Session", out.dir = getwd()) %>%
  kable()
```


# Results

## Pubmed report
```{r plot-pubmed-metatranscriptomics}
#| label: plot-pubmed-metatranscriptomics
#| fig-height: 5
#| fig-cap: '**Pubmed report for metatranscriptomic search (February 2024)**, number of paper that has metatranscriptomic word'

p0 <- pm %>%
  ggplot(aes(x = Year, y = Count)) +
  geom_col(fill = "#0072B2", color = "black") +  # Blue with black border
  labs(title = "Increase in Metatranscriptomics Publications Over Time",
       x = "Year", 
       y = "Number of Publications") +
  theme_minimal(base_size = 14)


pdf("figure1.pdf", width = 10, height = 10)  # apre il "dispositivo grafico PDF"
p0
dev.off()  # chiude e salva il file

p0
```

## kimchi

### Statistics plot - quality of mapping reads

-   Nf-core/metatdenovo gets much higher mapped reads compared to kimchi (@tbl-kimchi-stats)
-   Higher gap in J7 while progressively the gap decreases, probably because the isolate community incresases as well

```{r tbl-kimchi-stats }
#| label: tbl-kimchi-stats
#| tbl-cap: '**Mapped reads comparison** Comparison of mapped reads between nf-core/metatdenovo and kimchi original paper. J7 has more gap than the rest of the samples, probably due the distribution of community among the samples'
#| warning: false

stats.og <- data.frame(sample = c("J7", "J13", "J18", "J25", "J29"),
                       n_trimmed = c(30535784, 35269344, 33894898, 31437973, 32562239),
                       n_non_contaminated = c(20198679, 16650892, 10820351, 4250132, 5793508),
                       n_mapped = c(425552, 5332645, 8442826, 1768222, 3869480),
                       database = c( "kimchi", "kimchi", "kimchi", "kimchi", "kimchi"))

final_table <- k.overall_stats %>%
  mutate( database = c( "metatdenovo", "metatdenovo", "metatdenovo", "metatdenovo", "metatdenovo")) %>%
  rename( n_mapped = idxs_n_mapped) %>%
  select(sample, n_trimmed, n_non_contaminated, n_mapped, database) %>%
  rbind(stats.og) %>%
  mutate( total.reads = c(58133534, 70957538, 66160090, 61382647, 63560382,
                          58133534, 70957538, 66160090, 61382647, 63560382)) %>%
  relocate( database, sample, total.reads, n_trimmed, n_non_contaminated, n_mapped)

write.csv(final_table, "table_for_paper.csv", row.names = FALSE)

final_table %>%
  kable()
```

```{r table-kimchi-stats }
#| label: plot-kimchi-stats
#| tbl-cap: '**Mapped reads comparison** Comparison of mapped reads between nf-core/metatdenovo and kimchi original paper. J7 has more gap than the rest of the samples, probably due the distribution of community among the samples'

k.ordered.plots <- c("p_mapped_to_NC", "p_gtdb_to_NC", "p_FC_to_NC", "p_gtdb_to_mapped", "p_FC_to_mapped", "p_gtdb_to_FC")

k.overall_stats %>%
  mutate( database = c( "metatdenovo", "metatdenovo", "metatdenovo", "metatdenovo", "metatdenovo")) %>%
  rename( n_mapped = idxs_n_mapped) %>%
  mutate( total.reads = c(58133534, 70957538, 66160090, 61382647, 63560382)) %>% 
  group_by(sample) %>%
  mutate( p_mapped_to_NC = (n_mapped*100)/n_non_contaminated,
          p_FC_to_NC     = (n_feature_count*100)/n_non_contaminated,
          p_FC_to_mapped = (n_feature_count*100)/n_mapped,
          p_gtdb_to_NC   = (gtdb_eukulele_n_counts*100)/n_non_contaminated,
          p_gtdb_to_mapped = (gtdb_eukulele_n_counts*100)/n_mapped,
          p_gtdb_to_FC   = (gtdb_eukulele_n_counts*100)/n_feature_count
          ) %>%
    pivot_longer(p_mapped_to_NC:p_gtdb_to_FC, names_to = 'var', values_to = 'percentage') %>%
  ggplot(aes(x = factor(sample, levels = k.rank.sample), y = percentage)) +
  ylim(0,100) +
  scale_fill_manual(values = k.color.taxon) +
  facet_wrap(~factor(var, levels = k.ordered.plots), ncol = 2, drop = FALSE) +
  geom_col(width= .5, colour="black") +
  labs(title = "", x = "", y = "Mapped reads (%)") +
  coord_flip()
```

### Taxonomical annotation - comparison with original data

-   similar taxonomy annotation obtained for the species present in the original paper
-   our pipeline can show other groups of bacteria present in the samples
-   gtdb is a good database for this type of analysis

```{r taxonomoical annotation table - tpms, echo = FALSE, warning = FALSE}
#| fig-height: 10
#| warning: false
# build a table from original paper
taxonomy.op <- data.frame( sample = c(rep("J7",6), rep("J13",6), rep("J18",6),
                                      rep("J25",6), rep("J29",6)),
                           species = rep(c("Leuconostoc_carnosum", "Leuconostoc_gelidum",
                                        "Leuconostoc_inhae", "Weissella_koreensis",
                                        "Latilactobacillus_sakei", "Leuconostoc_mesenteroides"),5),
                           sum_tpm = c(0.089, 0.06, 0.04, 0.04, 0.05, 0.72,
                                          0.07, 0.14, 0.02, 0.20, 0.27, 0.30,
                                          0.02, 0.15, 0.05, 0.44, 0.22, 0.12,
                                          0.08, 0.04, 0.01, 0.64, 0.05, 0.18,
                                          0.015, 0.02, 0.005, 0.36, 0.45, 0.15),
                           sum_perc = c(0.18,0.12,0.08,0.08,0.1,1.44,
                                            2.24,4.48,0.64,6.4,8.64,9.6,
                                            1.56,11.7,3.9,34.32,17.16,9.36,
                                            3.36,1.68,0.42,26.88,2.1,7.56,
                                            1,1.34,0.33,24.12,30.15,10.05),
                           data = rep("kimchi",30) )


# combine the gtdb and phylodb taxonomy and keep only the best pid, if the pid is the same choose gtdb
k.taxonomy <- rbind(eukulele %>%
                filter( project == 'kimchi' ) %>%
                mutate_all(as.character) %>%
                mutate(
                  dplyr::across(
                  .cols = everything(), 
                  .fns = ~ dplyr::if_else(stringr::str_detect(.x, "NA uncl."), "uncl.", .x))) %>%
                select(-project, -run, -assembly, -assembler, -orf_caller, -bbnorm, -minlen, -taxon_rank) %>%
                as_tibble(), k.taxonomy.euk %>%
                  mutate( db = 'phylodb') %>%
                  relocate(db, .before = orf) %>%
                  select(-counts)) %>%
        group_by(orf) %>%
  filter( max_pid == max(max_pid))
  
k1 <- k.taxonomy %>%
  group_by(orf, max_pid) %>%
  filter(n() > 1 & all(db == 'gtdb_prior_to_2024019'))
k2 <- k.taxonomy %>%
  filter(!duplicated(orf))
k.tax.final <- bind_rows(k1, k2) %>%
  distinct(orf, .keep_all = TRUE)

# filter counts and change samples name
# create a variable with both taxonomy annotations
k.counts.tax.phylodb.gtdb <- counts %>%
  filter( project == 'kimchi') %>%
  mutate( sample = str_replace(sample, "SRR443366", "J7"),
          sample = str_replace(sample, "SRR443367", "J13"),
          sample = str_replace(sample, "SRR443368", "J18"),
          sample = str_replace(sample, "SRR443369", "J25"),
          sample = str_replace(sample, "SRR443370", "J29")) %>%
  left_join( (bind_rows(k1, k2) %>%
                distinct(orf, .keep_all = TRUE)), by = 'orf') %>%
  mutate( species = replace(species, species == 'Latilactobacillus_sakei_A', 'Latilactobacillus_sakei')) %>%
  mutate( order = replace(order, order == ' Lactobacillales', 'Lactobacillales'))

# create a variable with only gtdb taxonomy
k.counts.tax <- counts %>%
  filter( project == 'kimchi') %>%
  mutate( sample = str_replace(sample, "SRR443366", "J7"),
          sample = str_replace(sample, "SRR443367", "J13"),
          sample = str_replace(sample, "SRR443368", "J18"),
          sample = str_replace(sample, "SRR443369", "J25"),
          sample = str_replace(sample, "SRR443370", "J29")) %>%
  left_join( eukulele %>%
                filter( project == 'kimchi' ) %>%
                mutate_all(as.character) %>%
                mutate(
                  dplyr::across(
                  .cols = everything(), 
                  .fns = ~ dplyr::if_else(stringr::str_detect(.x, "NA uncl."), "uncl.", .x))) %>%
                select(-project, -run, -assembly, -assembler, -orf_caller, -bbnorm, -minlen, -taxon_rank) %>%
                as_tibble(), by = 'orf') %>%
  mutate( species = replace(species, species == 'Latilactobacillus_sakei_A', 'Latilactobacillus_sakei'),
          species = replace(species, species == 'Leuconostoc_mesenteroides_B', 'Latilactobacillus_sakei'))

# substitute NAs with 'uncl.'
k.counts.tax[is.na(k.counts.tax)] <- 'uncl.'

# bar plot with species found in the original paper
p0 <- k.counts.tax %>%
  mutate(project = ifelse(project == "kimchi", "Jung et al.", project)) %>%
  filter( species == "Leuconostoc_mesenteroides" |
            species == "Leuconostoc_carnosum" |
  species == "Leuconostoc_gelidum" | species == "Weissella_koreensis" |
  species == "Latilactobacillus_sakei" | species == "Leuconostoc_inhae") %>%
  group_by(sample, species) %>%
  summarise(sum_tpm = sum(tpm), .groups = 'drop') %>%
  mutate( data = "metatdenovo") %>%
  rbind( taxonomy.op[,-4]) %>%
  mutate(data = ifelse(data == "kimchi", "Jung et al.", data)) %>%
  ggplot(aes(x = factor(sample, levels = k.rank.sample), y = sum_tpm,
             fill = factor(species, level = k.rank.taxa))) +
  scale_fill_manual(values = k.color.taxon) +
  theme(legend.position = 'none') +
  geom_col(width= .5, colour="black", position = 'fill') +
  facet_wrap(~data) +
  labs(x = "", fill = "Species", y = "RE (%)")

p0
```

### Taxonomical annotation - proportion mapped reads

-   At species level nf-core/metatdenovo is following the same pattern as kimchi (@fig-kimchi-tax-annot A-B)
-   the proportion of mapped reads on at species level is different, higher in Kimchi. probably for less specificity (C-D)

```{r unused code for mapped proportion }
# metatdenovo output proportion mapped reads (FC vs gtdb)
j7 <- k.counts.tax %>%
  filter(sample == 'J7') %>%
  mutate(percentage = (count*100/13904698))

j13 <- k.counts.tax %>%
  filter(sample == 'J13') %>%
  mutate(percentage = (count*100/14985451))

j18 <- k.counts.tax %>%
  filter(sample == 'J18') %>%
  mutate(percentage = (count*100/10982168))

j25 <- k.counts.tax %>%
  filter(sample == 'J25') %>%
  mutate(percentage = (count*100/4065160))

j29 <- k.counts.tax %>%
  filter(sample == 'J29') %>%
  mutate(percentage = (count*100/6025743))

selected.species.nc.rds <- rbind(j7,j13,j18,j25,j29) %>%
  filter( species == "Leuconostoc_mesenteroides" | species == "Leuconostoc_carnosum" |
            species == "Leuconostoc_gelidum" | species == "Weissella_koreensis" | species == "Latilactobacillus_sakei" | species == "Leuconostoc_inhae") %>%
  group_by(sample, species) %>%
  summarise(sum_perc = sum(percentage), .groups = 'drop') %>%
  rename( taxon = 'species')

selected.other.nc.rds <- rbind(j7,j13,j18,j25,j29) %>%
  group_by(sample) %>%
  filter( species != "Leuconostoc_mesenteroides" & species != "Leuconostoc_carnosum" &
            species != "Leuconostoc_gelidum" & species != "Weissella_koreensis" & species != "Latilactobacillus_sakei" & species != "Leuconostoc_inhae") %>%
  mutate( taxon = 'Others') %>%
  group_by(sample, taxon) %>%
  summarise(sum_perc = sum(percentage), .groups = 'drop')

top.other.nc.rds <- selected.other.nc.rds %>%
  group_by(taxon, sample) %>%
  summarise(perc = sum(sum_perc),  .groups = 'drop_last') %>%
  summarise(mean_perc = mean(perc), .groups = 'drop') %>%
  slice_max(mean_perc, n = 5) %>%
  na.omit()

p1  <- selected.other.nc.rds %>%
  inner_join(top.other.nc.rds, by = 'taxon') %>%
  select(-mean_perc) %>%
  rbind(selected.species.nc.rds) %>%
  rename( species = 'taxon') %>%
  mutate( data = 'metatdenovo') %>%
  rbind(taxonomy.op[,-3]) %>%
  ggplot(aes(x = factor(sample, levels = k.rank.sample), y = sum_perc, fill = factor(species, levels = k.rank.p1))) +
  ylim(0,100) +
  scale_fill_manual(values = k.color.taxon) +
  theme(legend.text = element_text(size = 6),
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.position = c(0.5, 0.87),  # Move the legend slightly to the left
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.y = element_text(size = 10)) +
  guides(fill = guide_legend(nrow = 3))  +
  geom_col(width= .5, colour="black", size = 0.5) +
  facet_wrap(~data) +
  labs(x = "", fill = "taxon", y = "Mapped reads (%)")
p1
```

```{r fig-kimchi-mapped-proportion }
#| label: fig-kimchi-mapped-proportion
#| warning: false
#| fig-cap: '**Mapped proportion**'
#| fig-height: 10


# Define reusable variables/functions
target_species <- c(
  "Leuconostoc_mesenteroides", "Leuconostoc_carnosum",
  "Leuconostoc_gelidum", "Weissella_koreensis",
  "Latilactobacillus_sakei", "Leuconostoc_inhae"
)

sample_totals <- tribble(
  ~sample,  ~total_reads,
  "J7",     13904698,
  "J13",    14985451,
  "J18",    10982168,
  "J25",    4065160,
  "J29",    6025743
)

# Calculate percentages and categorize taxa
taxa_percentages <- k.counts.tax %>%
  mutate(project = ifelse(project == "kimchi", "Jung et al.", project)) %>%
  left_join(sample_totals, by = "sample") %>%
  mutate(percentage = (count * 100) / total_reads) %>%
  mutate(
    taxon = case_when(
      species %in% target_species ~ species,
      TRUE ~ "Others"
    )
  )

# Summarize percentages for target species and top "Others"
plot_data <- taxa_percentages %>%
  group_by(sample, taxon) %>%
  summarise(sum_perc = sum(percentage), .groups = "drop") %>%
  # Identify top 5 "Others" (if needed)
  mutate(
    taxon = if_else(
      taxon == "Others" & sum_perc < min(
        filter(., taxon %in% target_species) %>% pull(sum_perc)
      ),
      "Other low-abundance",
      taxon
    )
  ) %>%
  rename(species = taxon) %>%
  mutate(data = "metatdenovo") %>%
  bind_rows(taxonomy.op[, -3])  # Combine with external data

# Mapped reads (%)
p1 <- plot_data %>%
  mutate(data = ifelse(data == "kimchi", "Jung et al.", data)) %>%
  ggplot(aes(
    x = factor(sample, levels = k.rank.sample),
    y = sum_perc,
    fill = factor(species, levels = k.rank.p1)
  )) +
  geom_col(width = 0.5, colour = "black", size = 0.5) +
  scale_fill_manual(values = k.color.taxon) +
  labs(x = "", y = "Mapped reads (%)", fill = "Taxon") +
  ylim(0, 100) +
  facet_wrap(~data) +
  theme(
    #legend.text = element_text(size = 6),
    legend.title = element_blank(),
    legend.key.size = unit(0.3, "cm"),
    legend.text = element_text(size = 7),
    legend.position = 'top',
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.y = element_text(size = 10)
  ) +
  guides(fill = guide_legend(nrow = 3))

# Top orders (TPM)
p2_data <- k.counts.tax.phylodb.gtdb %>%
  mutate(
    project = ifelse(project == "kimchi", "Jung et al.", project),
    order = case_when(
      domain == "Viruses" ~ "Viruses",
      phylum == " Archaeplastida" ~ "Plants",
      order == "Bacteria uncl. uncl. uncl." ~ "Bacteria",
      order == "Gammaproteobacteria uncl." ~ "Gammaproteobacteria",
      TRUE ~ order
    )
  ) %>%
  group_by(order, sample) %>%
  summarise(sum_tpm = sum(tpm), .groups = "drop")

top_orders <- p2_data %>%
  filter(order != "uncl.") %>%
  group_by(order) %>%
  summarise(mean_tpm = mean(sum_tpm), .groups = "drop") %>%
  slice_max(mean_tpm, n = 15)

p2 <- p2_data %>%
  inner_join(top_orders, by = "order") %>%
  ggplot(aes(
    x = factor(sample, levels = k.rank.sample),
    y = sum_tpm,
    fill = factor(order, levels = k.rank.order)
  )) +
  geom_col(width = 0.5, colour = "black", size = 0.3) +
  scale_fill_manual(values = k.color.taxon) +
  labs(x = "", y = "TPM", fill = "Major taxa") +
  theme(
    legend.text = element_text(size = 7),
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.position = c(0.4, 1.2),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.y = element_text(size = 10)
  ) +
  guides(fill = guide_legend(nrow = 3))
p2
```

### Functional annotation - heatmap

-   The heatmap shows how many genes related to fermentative processeses. (@fig-kimchi-heatmap)
-   Some processes are specific for the Genus while others are present in all Genus

```{r fig-kimchi-heatmap}
#| label: fig-kimchi-heatmap
#| warning: false
#| fig-cap: '**Heatmap**'
#| fig-height: 10

# Code for enzymes
# 1, Glucokinase; 2, Glucose-6-phosphate isomerase; 3, 6-Phosphofructokinase; 
# 4, Fructose-bisphosphate aldolase; 5, Triose-phosphate isomerase; 
# 6, Glyceraldehyde-3-phosphate dehydrogenase; 7, Phosphoglycerate kinase; 
# 8, Phosphoglycerate mutase; 9, Enolase; 10, Pyruvate kinase; 11, Lactate dehydrogenase;
# 12, Glucose-6-phosphate 1-dehydrogensase; 13, 6-Phosphogluconate dehydrogenase;
# 14, Ribulose-5-phosphate 3-epimerase; 15, Xylulose 5-phosphate phosphoketolase;
# 16, Phosphotransacetylase; 17, Bifunctional acetaldehyde-CoA/Alcohol dehydrogenase; 
# 18, Acetate kinase; 19, Mannitol dehydrogenase; 20, Fructokinase.

filtered_genes <- k.counts.tax %>%
  select( -minlen) %>%
  left_join(k.prokkas, by = join_by(orf, start, end, strand)) %>%
  left_join(k.eggnogs %>% filter(project == 'kimchi'), by = join_by(project, run, assembler, orf_caller, bbnorm, orf, assembly)) %>%
  mutate( species = replace(species, species == 'Latilactobacillus_sakei_A', 'Latilactobacillus_sakei')) %>%
  mutate(product = replace(product, pfams == 'Enolase_C,Enolase_N', 'Enolase')) %>%
  mutate(product = replace(product, pfams == 'His_Phos_1', 'phosphoglycerate mutase')) %>%
  mutate(product = replace(product, 
                              pfams == 'Mannitol_dh,Mannitol_dh_C', 'Mannitol dehydrogenase')) %>%
  mutate(product = replace(product, product == 'Enolase 2', 'Enolase')) %>%
  mutate(product = replace(product, 
                              product == 'L-lactate dehydrogenase', 'lactate dehydrogenase')) %>%
  mutate(product = replace(product, 
                              product == 'D-lactate dehydrogenase', 'lactate dehydrogenase')) %>%
  mutate(product = replace(product, 
                              product == 'Quinone-dependent D-lactate dehydrogenase',
                              'lactate dehydrogenase')) %>%
  mutate(product = replace(product, preferred_name == 'fba', 'Aldolase')) %>%
  mutate(product = replace(product, 
                              description == 'Triosephosphate isomerase', 'Triosephosphate isomerase')) %>%
  mutate(product = replace(product, description == 'Belongs to the glyceraldehyde-3-phosphate dehydrogenase family', 'Glyceraldehyde-3-phosphate dehydrogenase')) %>%
  mutate(product = replace(product, description == 'Belongs to the phosphoglycerate kinase family', 'Phosphoglycerate kinase')) %>%
  mutate(product = replace(product, preferred_name == 'glcK', 'Glucokinase')) %>%
  mutate(product = replace(product, description == 'L-lactate dehydrogenase (FMN-dependent) and related alpha-hydroxy acid dehydrogenases', 'lactate dehydrogenase')) %>%
  mutate(product = replace(product, preferred_name == 'gnd', '6-Phosphogluconate dehydrogenase')) %>%
  mutate(product = replace(product, pfams == '6PGD,NAD_binding_2', '6-Phosphogluconate dehydrogenase')) %>%
  mutate(product = replace(product, description == 'Phosphoketolase', 'Phosphoketolase')) %>%
  mutate(product = replace(product, description == 'D-xylulose 5-phosphate/D-fructose 6-phosphate phosphoketolase', 'Phosphoketolase')) %>%
  mutate(product = replace(product, product == 'Putative fructokinase', 'Fructokinase')) %>%
  mutate(product = replace(product, description == '6-phosphate phosphoketolase', 'Phosphoketolase')) %>%
  mutate(product = replace(product, preferred_name == 'zwf', 'G6p1dh')) %>%
  mutate(product = replace(product, product == "Glucokinase", "glK"),
         product = replace(product, product == "Glucose-6-phosphate isomerase", "	
pgi"),
         product = replace(product, product == "ATP-dependent 6-phosphofructokinase", "pfkA"),
         product = replace(product, product == "Aldolase", "fba"),
         product = replace(product, product == "Triosephosphate isomerase", "tpiS"),
         product = replace(product, product == "Glyceraldehyde-3-phosphate dehydrogenase", "gap"),
         product = replace(product, product == "Phosphoglycerate kinase", "pgk"),
         product = replace(product, product == "phosphoglycerate mutase", "pgm"),
         product = replace(product, product == "Enolase", "eno"),
         product = replace(product, product == "Pyruvate kinase", "pyk"),
         product = replace(product, product == "lactate dehydrogenase", "lctd"),
         product = replace(product, product == "G6p1dh", "G6p1dh"),
         product = replace(product, product == "6-Phosphogluconate dehydrogenase", 	
"gntZ"),
         product = replace(product, product == "Phosphoketolase", "	
xpkA"),
         product = replace(product, product == "Mannitol dehydrogenase", "	
uxuB"),
         product = replace(product, product == "Fructokinase", "	
scrK"),
         product = replace(product, product == "Acetate kinase", "actk")
         ) %>%
  filter( product %in% 
            c("glK", "pgi", "pfkA",
              "fba", "tpiS", "gap",
              "pgk", "pgm", "eno", "pyk",
              "lctd", "G6p1dh", "gntZ",
              "xpkA", "uxuB", "scrK", "actk")) %>%
  filter( genus %in% c("Latilactobacillus", "Leuconostoc", "Weissella")) %>%
  mutate(genus = fct_relevel(genus, "Leuconostoc", "Latilactobacillus",
                             "Weissella"))

# set colors and contants for heatmap
k.custom_colors <- c("blue", "light blue", "green", "yellow", "red")
breaks <- c(1, 10, 4000, 8000)
color_thresholds <- c(0, 100, 1000, 10000, 15000, Inf)
color_labels <- c("0", "100", "1000", "10000", ">15000")

p3 <- filtered_genes %>%
  select(-tpm) %>%
  mutate( r = count/length) %>%
  group_by(sample) %>% 
  mutate(tpm = r/sum(r) * 1e6) %>% 
  ungroup() %>%
  select(-r) %>%
  group_by(genus, sample, product) %>%
  summarise(sum_tpm = sum(tpm), .groups = 'drop') %>%
  mutate(ColorCategory = cut(sum_tpm, breaks = color_thresholds, labels = color_labels, include.lowest = TRUE)) %>%
  ggplot(aes(x = factor(sample, level = k.rank.sample), y = factor(product, level = k.rank.genes),
             fill = ColorCategory)) +
  geom_tile(colour="black", size = 0.1) +
  scale_fill_manual(
     values = c(
    "0" = "#999999",     # grey - neutral low value
    "100" = "#E69F00",   # orange - distinguishable and soft
    "1000" = "#56B4E9",  # sky blue - high contrast
    "10000" = "#009E73", # teal green - calm and clear
    ">15000" = "#D55E00" # reddish-brown - attention without being flashy
    ),
    breaks = color_labels,
    labels = color_labels,
    guide = "legend"
  ) +
  facet_wrap(~ genus) +
  labs(x = "", y = "") +
  theme(
    axis.text.x = element_text(angle = -90, size = 8, hjust = 1),
    panel.grid = element_blank(),
    legend.position = 'right',
    legend.title = element_blank(),
    #legend.key.size = unit(0.1, "cm")
    ) +
  coord_fixed(ratio = 0.9)
 
px <- (p1 + p0) /p3 + plot_annotation(tag_levels = 'A')


pdf("figure.pdf", width = 10, height = 10)  # apre il "dispositivo grafico PDF"
px
dev.off()  # chiude e salva il file

px
```


